# Texas 2016 US Presidential Campaign Contribution Analysis 
## by Patrick Bloomingdale
This project is part of Udacity's Data Analyst Nanodegree <br>
**Term:** 2 <br>
**Project:** 2 <br>
**Due Date:** November 13, 2018

========================================================

## Introduction

For this project, I will conduct an exploratory data analysis (EDA) and create 
an RMD file that explores the variables, structure, patterns, oddities, and 
underlying relationships the following dataset: <br>
> [**2016 Presidential Campaign Finance Data**](https://classic.fec.gov/disclosurep/PDownload.do)
<br>

### Dataset Considerations

**CONTRIBUTION LIMITS**<br>
[**Contribution limits for 2015-2016**](https://www.fec.gov/updates/contribution-limits-for-2015-2016/) <br>
Under the Federal Election Campaign Act (the Act), the limits on contributions 
made by persons to candidates (increased to $2,700 per election, per candidate)
(52 U.S.C. ยง 30116(a)(1)(A)). The per-election limits on contributions to 
candidates are in effect for the two-year election cycle beginning the day after 
the general election and ending on the date of the next general election 
(November 5, 2014 - November 8, 2016). All other contribution limits are in 
effect for the two-calendar-year period beginning on January 1, 2015 and ending 
on December 31, 2016.
<br>

[**2016 PRESIDENTIAL ELECTION RESULTS FOR THE UNITED STATES**](https://www.270towin.com/2016_Election/) <br>
The United States presidential election of 2016 was held on Tuesday, 
November 8, 2016. The Republican canidate Donald Trump defeated the Democratic 
canidate Hillary Clinton. <br>

The table below represents the results from the 2016 presidential election. 
The only canidates included in this table are the ones that received the 
parties nomination.

| Candidate           | Party         | Electoral Votes | Popular Votes |
| ------------------- | ------------- | ---------------:| -------------:|
| Donald J. Trump     | Republican    | 304   		    | 62,980,160    |
| Hillary R. Clinton  | Democratic    | 227   		    | 65,845,063    |
| Gary Johnson        | Libertarian   | 0   		    | 4,488,931     |
| Jill Stein  		  | Green    	  | 0  			    | 1,457,050     |
| Evan McMullin  	  | Independent   | 0   		    | 728,830       |

<br>
[**2016 PRESIDENTIAL ELECTION RESULTS FOR THE STATE OF TEXAS**](https://www.politico.com/2016-election/results/map/president/texas/) <br>
The table below represents Texas' results from the 2016 presidential election. 
The only canidates included in this table are the ones that received the 
parties nomination.

| Candidate           | Party         | Popular Votes  | Percentage |
| ------------------- | ------------- | --------------:| ----------:|
| Donald J. Trump     | Republican    | 4,681,590      | 52.6       |
| Hillary R. Clinton  | Democratic    | 3,867,816      | 43.4       |
| Gary Johnson        | Libertarian   | 282,524   	   | 3.2        |
| Jill Stein  		  | Green    	  | 71,307 		   | 0.8        |
| Evan McMullin*  	  | Independent   | na   		   | na         |
*Dataset did not contain any votes for canidate

**Party Affiliation** <br>
To ensure I have the corrected canidate associated with the correct party, 
I downloaded a list of all the canidates with their party affiliation from the 
[Federal Election Commission](https://www.fec.gov/data/advanced/?tab=candidates)
, then compared it to the list of canidate names to assign a party affiliation.

**Initial Exploratory Questions** <br>
1. Is there a relationship between campaign contributions and party affiliation? <br>
2. What is the total ammount of campaign contributions per canidate? <br>
3. Campaign contributions and gender? <br>
4. Campaign contributions and occupation? <br>

**ADDITIONAL RESOURCES** <br>
[**Democratic Convention**](http://www.thegreenpapers.com/P16/D)<br>
[**Democratic Convention - Texas**](http://www.thegreenpapers.com/P16/TX-D#TX-D-STATEWIDE)<br>
[**Republican Convention**](http://www.thegreenpapers.com/P16/R)<br>
[**Republican Convention - Texas**](http://www.thegreenpapers.com/P16/TX-R#TX-R-STATEWIDE)<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
#set knitr options for figure sizes
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)

#load packages
library(ggplot2)
library(gender)
#install_github("ropensci/genderdata")
library(genderdata)
library(gridExtra)
library(dplyr)
#install.packages("stringi", repos="http://cran.rstudio.com/", dependencies=TRUE)
library(remotes)
library(zipcode)
library(aod)
library(scales)
library(tidyr)

options(scipen = 999)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
tx <- read.csv("P00000001-TX.csv", row.names = NULL)
colnames(tx) <- c(colnames(tx)[-1],"X")
tx$X <- NULL
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# drop columns not going to use
drops <- c("cmte_id","receipt_desc", "memo_cd", "memo_text")
tx <- tx[ , !(names(tx) %in% drops)]
```

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}
#type of object, observations, variables
str(tx)
```

**FINDINGS** <br>
The dataset contains the following: <br>
 - 548,396 observations (rows) and 14 variables (columns)<br>
 - 25 Canidates<br>
 - 129,927 Contributors<br>
 
##  1. Contribution Amounts (Univariate) 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_1a}
#contribution amount histogram before removing negative values
ggplot(aes(x = contb_receipt_amt), 
       data = tx) +
  geom_histogram(bins = 50) +
  xlab('Contribution Amount') +
  ggtitle('Number of Contribution Amounts Received - Histogram')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#determine the number of negative contributions 
sum(tx$contb_receipt_amt < 0)
```

**FINDINGS**<br>
The histogram above shows that the dataset has negative contribution amounts. I 
am going to remove the 12,084 values less than zero to get a better idea of what 
the contributions look like. Then I am going to use a boxplot to look for any 
outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#remove observations that have contribution amounts less than zero
tx <- tx[tx$contb_receipt_amt > 0,]
#check to see if removed
sum(tx$contb_receipt_amt < 0)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_1b}
#contribution amount boxplot with negative values removed
#showing outliers
ggplot(aes(x = 1, y = contb_receipt_amt), 
       data = tx) +
  geom_boxplot() +
  ggtitle('Contribution Amounts Received - Boxplot')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#determine the number of contributions greater than 2700
sum(tx$contb_receipt_amt >= 2700)
```

**FINDINGS**<br>
The boxplot above shows that their are lots of outliers making it hard to get a 
detailed analysis of the contributions.  As stated earlier in the `Introduction` 
section, the maximum amount an individual can contribute is `$2,700`. I am going
to remove the 10,093 contributions that are greater than the $2,700 because they 
should have been refunded.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#remove observations that excede the limit of 2,700
tx <- tx[tx$contb_receipt_amt <= 2700,]
#check to see if removed
sum(tx$contb_receipt_amt > 2700)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_1c}
#summary of contb_receipt_amt
summary(tx$contb_receipt_amt)

#contribution amount histogram using log10
ggplot(aes(x = contb_receipt_amt), 
       data = tx) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  ggtitle('Contribution Amount (Log10) - Histogram') +
  xlab('Contribution Amount') +
  ylab('Number of Contributions')
```

**FINDINGS**<br>
By using scale_x_log10() I was able to get a better look at the distribution of 
contributions. <br> 
 - More than half the contributions were less than $40.00 (Median) <br>
 - Approximately 25% of the data values are less than or equal to $25.00 
 (First Quartile) <br>
 - Approximately 25% of the data values are less than or equal to $100.00 
 (Third Quartile) <br>
 
I have determined that I need to add more variables and clean up others to do a 
more detailed analysis.

##  2. Contributions to Party (Univariate) 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_2}
# create party variables
democrat <- c("Clinton, Hillary Rodham", "Lessig, Lawrence", 
              "O'Malley, Martin Joseph", "Sanders, Bernard", 
              "Webb, James Henry Jr.")
tx$party <- ifelse(tx$cand_nm %in% democrat, "Democrat", "Republican")
tx$party[tx$cand_nm %in% c("Stein, Jill", "McMullin, Evan", 
                           "Johnson, Gary")] <- 'Other'

# Create party dataset
party_aff <- group_by(tx, party)
tx.contr_party <- summarize(party_aff,
                               sum_party = sum(contb_receipt_amt),
                               number_of_candidate = length(unique(cand_id)), 
                               mean_party = sum_party/number_of_candidate, 
                               n = n())

#total contributions
sum(tx.contr_party$n)

# contr_party table
tx.contr_party

#number of contributios to party bar chart
ggplot(aes(x = party, y = n), 
       data = tx.contr_party) +
  geom_bar(stat = 'identity') +
  geom_text(stat = 'identity', aes(label = n),
            data = tx.contr_party, vjust = -0.4) +
  xlab('Party') +
  ylab('Number of Contributions') +
  ggtitle('Number of Contributions to Party') 
```

**FINDINGS**<br>
 - The total number of contributions made was approximately 517,647<br> 
 - The Democratic Party had the most contributions at 271,863<br>
 - The Republican Party had the second most contributions 243,948<br> 
 - All other parties had only 1,836 contributions<br>
 
##  3. Contributions to Canidate (Univariate) 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_3}
#table of canidates
table(tx$cand_nm)

#number of contributions to canidate bar chart
ggplot(aes(x = cand_nm), 
       data = tx) + 
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab('candidate') +
  ylab('Number of Contributions') +
  ggtitle('Number of Contributions by Candidate')
```

**FINDINGS**
<br>
 - Hillary Clinton had the most contributions at 194,289 <br>
 - Ted Cruz had the second most contributions at 127,925 <br>
 - Bernard Sanders had the third most contributions at 77,278 <br>
 - Donald Trum had the forth most contributions at 73,443 <br>
 - Benjamin Carson has the fifth most contributions at 22,471 <br>

##  4. Contributions by Gender (Univariate)

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_4}
# get contributor's first name to determine gender 
tx$contbr_first_nm <- sub(" .*", "", sub(".*, ", "", tx$contbr_nm))

# create gender dataset
gender_df <- gender(tx$contbr_first_nm, method = 'ssa', c(1920, 1997),
                       countries = 'United States')
gender_df <- unique(gender_df)
names(gender_df)[1] <- 'contbr_first_nm'
tx <- inner_join(tx, gender_df, by = 'contbr_first_nm')

#create tx.contr_gen
gender_group <- group_by(tx, gender)
tx.contr_gen <- summarize(gender_group, 
                             sum_gen = sum(contb_receipt_amt),
                             n_gen = n())
tx.contr_gen

#number of contributions by gender bar chart
ggplot(aes(x = gender, y = n_gen), 
       data = tx.contr_gen, vjust = -0.4) +
  geom_bar(stat = 'identity') +
  geom_text(aes(label = n_gen), stat = 'identity', 
            data = tx.contr_gen, vjust = -0.4) +
  xlab('Gender') +
  ylab('Number of Contributions') +
  ggtitle('Number of Contributions by Gender') 
```

**FINDINGS**<br>
Of the total 517,647 contributions: <br>
 - Females accounted for approximately 243,640 (47%) of contributions<br>
 - Males accounted for approximately 274,007 (53%) of contributions<br>

##  5. Contributions by Occupation (Univariate)

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_5}
# create occupation dataset
occ_group <- group_by(tx, contbr_occupation)
tx.contr_occ <- summarize(occ_group, 
                              sum_occu = sum(contb_receipt_amt), 
                              mean_occu = mean(contb_receipt_amt), 
                              n = n())
tx.contr_occ <- subset(tx.contr_occ, contbr_occupation != "INFORMATION REQUESTED")
tx.contr_occ <- head(arrange(tx.contr_occ,desc(n)), n = 5)
tx.contr_occ$contbr_occupation <- ordered(tx.contr_occ$contbr_occupation, 
                                              levels = c("RETIRED", 
                                                         "NOT EMPLOYED", 
                                                         "ATTORNEY", 
                                                         "HOMEMAKER", 
                                                         "ENGINEER"))

contr_occ <- c("RETIRED", "NOT EMPLOYED", "ATTORNEY", "HOMEMAKER", "ENGINEER")

#tx.contr_occ table
tx.contr_occ

#top five contributors by occupation
ggplot(aes(x = contbr_occupation, y = n), 
       data = tx.contr_occ) +
  geom_bar(stat = 'identity') +
  xlab('Occupations') +
  ylab('Number of Contributors') +
  ggtitle('Top Five Contributors by Occupation') 
```

**FINDINGS**<br>
Since there were 16,992 different occupations listed in the data set I decided 
to only analyze the top five contributor occupations:<br>
 - Retired at 135,939<br>
 - Not Employed at 22,635<br>
 - Attorney at 13,516<br>
 - Homemaker at 10,358<br>
 - Engineer at 8,484<br>

# Univariate Analysis

### What is the structure of your dataset?
 - 548,396 observations (rows) and 14 variables (columns)<br>
 - 25 Canidates<br>
 - 129,927 Contributors<br>
 
### What is/are the main feature(s) of interest in your dataset?
 The maid features of the dataset I am interested in are
 `contribution amounts`(contb_receipt_amt), the `contributor` (contbr_nm), and
 the `canidate` (cand_nm).
 
### What other features in the dataset do you think will help support your 
### investigation into your feature(s) of interest?
Other features in the dataset that will help support my investigation include:<br>
 - Total Number of Contributions to Party<br> 
 - Number of Contributions to Candidate<br>
 - Number of Contributions by Gender<br>
 - Top Occupations by Number of Contributors<br>
 - Date Contributions were Made<br>

### Did you create any new variables from existing variables in the dataset?
**I created the following variables for a more detailed analysis:**<br>
-`party:` candidate party affilliation (Democrat, Republican, Other)<br>
-`gender:` contributor gender (male, female)<br>

### Of the features you investigated, were there any unusual distributions?
 - There were 10,093 contributions that were greater than the contribution 
 limit of $2,700 so I removed them.
 - There were 12,084 contributions that were negative values so I removed them.

### Did you perform any operations on the data to tidy, adjust, or change the 
### form of the data? 
**I removed the following variables:**<br>
-`receipt_desc`<br>
-`memo_cd`<br>
-`X`<br>
-`proportion_male`<br>
-`proportion_female`<br>
-`year_min`<br>
-`year_max`<br>

**I cleaned the following variables:**<br>
-`contb_receipt_dt:` now the date is in %d-%b-%y format<br>
-`contb_receipt_amt:` removed all negative contributions and contributions 
greater than the limit of $2700<br>

# Bivariate Plots Section

## 1. Contribution Amount and Party (Bivariate)

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_1a}
# contribution amount by party table
tx.contr_party

# contribution amount by party bar chart
ggplot(aes(x = party, y = sum_party/1000),
       data = tx.contr_party) +
  geom_bar(stat = 'identity') +
  geom_text(stat = 'identity', 
            aes(label = round(sum_party/1000)),
            data = tx.contr_party, vjust = -0.4) +
  xlab('Party') +
  ylab('Contribution Received (Thousands)') +
  ggtitle('Total Contribution Amount to Party') 
```

**FINDINGS** <br>
 - The Republican Party raised approximately $47 million through contributions<br>
 - The Democratic Party raised approximately $25 million through contributions<br>
 - All other parties raised approximately $44 thousand through contributions<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_1b}
#summary of party contrb amount
by(tx$contb_receipt_amt, tx$party, summary)

#Contribution Amount (Log10) to Party Boxplot
ggplot(aes(x = party, y = contb_receipt_amt),
       data = tx) +
  geom_boxplot() +
  scale_y_log10() +
  xlab('party') +
  ylab('Contribution Amount') +
  ggtitle('Contribution Amount (Log10) to Party Boxplot') 
```

**FINDINGS** <br>
The Republican Parties mean and median contributions were higher than the 
Democratic Parties. You can also see from the boxplot above that the other 
parties mean and median was higher than both Republicans and Democrats. 

## 2. Contribution Amount and Canidate (Bivariate)

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_2a}
# create candidate dataset
can_group <- group_by(tx, party, cand_nm)
tx.contr_can <- summarize(can_group, 
                             sum_can = sum(contb_receipt_amt), 
                             mean_can = mean(contb_receipt_amt),
                             n = n())
tx.contr_can <- arrange(tx.contr_can, sum_can)

# contr_by_can table
tx.contr_can

#Contribution Received by Candidate bar chart
ggplot(aes(x = cand_nm, y = sum_can/1000), 
       data = tx.contr_can) +
  geom_bar(stat = 'identity') +
  scale_y_continuous(limits = c(0, 23000)) + 
  xlab('Candidate') +
  ylab('Contribution Received (Thousands)') +
  ggtitle('Contribution Received by Candidate') +
  coord_flip()
```

**FINDINGS** <br>
The canidates that recieved the most contributions were: <br>
 - H. Clinton totaled approximately $21.8 million <br> 
 - T. Cruz totaled approximately $18.3 million <br> 
 - D. Trump totaled approximately $15.3 million <br> 
 - J. Bush totaled approximately $3.6 million <br> 
 - B. Sanders totaled approximately $3.4 million <br> 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_2b}
#summary of cand_nm contrb amount
by(tx$contb_receipt_amt, tx$cand_nm, summary)

#Contribution Amount to Candidate box plot
ggplot(aes(x = cand_nm, y = contb_receipt_amt), 
       data = tx) + 
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab('candidate') +
  ylab('Contribution Amount') +
  ggtitle('Contribution Amount to Candidate')
```

**FINDINGS**<br>
 - Canidates with the highest median contributions:
 > C. Christie, R. Perry, R. Santorum, J. Bush, and L. Graham
 - Canidates with the lowest median contributions:
 > B. Carson, H. Clinton, T. Cruz, R. Paul, and B. Sanders.
 - Both Clinton's and Sanders' outliers are almost a solid black line showing 
 that they received a lot of small contribution amounts.
 
## 3. Contribution Amount by Gender (Bivariate)

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_3a}
by(tx$contb_receipt_amt, tx$gender, summary)

#Contribution Amount by Gender Boxplot
ggplot(aes(x = gender, y = contb_receipt_amt),
       data = tx) +
  geom_boxplot() +
  xlab('gender') +
  ylab('Contribution Amount') +
  ggtitle('Contribution Amount by Gender Boxplot') +
  coord_cartesian(ylim = c(0, 100))
```

**Findings**
By looking at the box plot and summary statistics above you can see the 
following differences between female and males:
 - the median contribution of males at $50.00 was higher than females the median 
 contributions at $35.00.
 - males also had a higher average contribution amount of 160.96 compared to 
 females at $119.44.  This is a difference of $41.52.

## 4. Contribution Amount by Occupation (Bivariate)

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_4}
#tx.contr_occ table
tx.contr_occ

#top five contributors by occupation
ggplot(aes(x = contbr_occupation, y = sum_occu/1000), 
       data = tx.contr_occ) +
  geom_bar(stat = 'identity') +
  xlab('Occupations') +
  ylab('Contribution Amount (Thousands)') +
  ggtitle('Contribution Amount by Top 5 Occupations')
```

**FINDINGS** <br>
The occupation that contributed the highest amount was retired people 
contributing over $14.5 million. This was followed by: <br>
 - Attorney at $3,970,684<br>
 - Homemaker at $3,396,749<br>
 - Engineer at $1,178,076<br>
 - Not Employed at $1,115,696<br>
 
# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the 
### investigation. How did the feature(s) of interest vary with other features  
### in the dataset?

**Party Affiliation**<br>
The Republican Party raised the most money at $47 million through contributions.
> Donald Trump (Republican) won the 2016 Presidential election.
<br>

**Canidate** <br>
Even though Donald Trump was the Republican party nomination, Ted Cruz raised
more money in Texas.  This might be due to the fact that Ted Cruz was a Texas
Senator when the election took place.

**Contributor Gender** <br>
Males accounted for more contributions and contributed larger amounts than 
females.

**Contributor Occupation** <br>
Retired people accounted for the most contributors by occupation.

### Did you observe any interesting relationships between the other features 
### (not the main feature(s) of interest)? 
The one that stood out most was when looking at the relationship between the 
canidate and contribution amounts I found that had Hillary Clinton and Bernard
Sanders had a lot of outliers showing that they received a lot of small 
contribution amounts.

### What was the strongest relationship you found?
Males accounted for more contributions and contributed larger amounts than 
females.

# Multivariate Plots Section

## 1. Contribution amounts by Canidate and Date (Multivariate)

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plot_1}
# adjust contb_receipt_dt
tx$contb_receipt_dt <- as.Date(tx$contb_receipt_dt, format = "%d-%b-%y")

# create top_five_can
top_five_can <- c("Clinton, Hillary Rodham", "Cruz, Rafael Edward 'Ted'", 
                  "Trump, Donald J.", "Bush, Jeb", "Sanders, Bernard")

## create tx.top_five_can dataset with general election date filter
tx.top_five_can <- tx %>%
  filter(contb_receipt_dt >= as.Date("2014-11-05") &
           contb_receipt_dt <= as.Date("2016-11-08") & 
                                         cand_nm %in% top_five_can)%>%
  group_by(cand_nm, contb_receipt_dt) %>%
  summarize(sum_top_can = sum(contb_receipt_amt),
            n = n())

# Contribution Amount by Canidate, by Contribtution Date
ggplot(aes(x = contb_receipt_dt, y = sum_top_can/1000, color = cand_nm), 
       data = tx.top_five_can) +
  geom_jitter(alpha = 1/10) +
  geom_smooth(method = 'loess') +
  scale_y_continuous() +
  xlab('Date of Contribution') +
  ylab('Contribution Amount (Thousands)') +
  ggtitle('Contribution Amount by Canidate, by Contribtution Date')

# Contribution Amount by Canidate, by Contribtution Date
# added limits to y axis
ggplot(aes(x = contb_receipt_dt, y = sum_top_can/1000, color = cand_nm), 
       data = tx.top_five_can) +
  geom_jitter(alpha = 1/10) +
  geom_smooth(method = 'loess') +
  scale_y_continuous(limits = c(0, 150)) +
  xlab('Date of Contribution') +
  ylab('Contribution Amount (Thousands)') +
  ggtitle('Contribution Amount by Canidate, by Contribtution Date')
```

**FINDINGS** <br>
Based on the two plots above, Ted Cruz began raising money before the other 
canidates. Around December of 2015, Hillary Clinton's contributions began to 
increase, and soon after Bernard Sanders' contributions began to decline.  
Donald Trumps contributions began to increase around March of 2016, overtake
all other canidates around July 2017 (except for Hillary Clinton).

## 2. Contribution amounts by Occupation and Date (Multivariate)

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plot_2}
## create tx.contr_occ dataset with general election date filter
tx.contr_occ <- tx %>%
  filter(contbr_occupation %in% contr_occ)%>%
  group_by(contbr_occupation, contb_receipt_dt) %>%
  summarize(sum_contb_amt = sum(contb_receipt_amt),
            n = n())

# Contribution Amount by Top 5 Occupation, by Contribtution Date
ggplot(aes(x = contb_receipt_dt, y = sum_contb_amt/1000, 
                 color = contbr_occupation), 
             data = tx.contr_occ) +
  geom_jitter(alpha = 1/10) +
  geom_smooth(method = 'loess') +
  scale_y_continuous(limits = c(0, 50)) +
  xlab('Date of Contribution') +
  ylab('Contribution Amount (Thousands)') +
  ggtitle('Contribution Amount by Top 5 Occupation, by Contribtution Date')
```

**FINDINGS** <br>
**Retired** <br>
 - Most contributions began coming in towards the end of 2014<br>
 - Contributions spiked towards the end of 2015, leveled off, and then began to 
 increase towards the middle of 2016 <br>

**Attorney** <br>
 - Contributions stayed steady throughout the general election period, but began 
 to increase around July 2016 <br>

**Homemaker** <br>
 - Contributions spiked around July 2015, leveled off, and then began to decline 
 steadily starting around January 20 <br>

**Engineer** <br>
 - Contributions stayed steady throughout the general election.<br>

**Not Employed** <br>
 - Contributions spiked and began to steadily decline starting around March 
 2016 <br>
 
# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the
### investigation. Were there features that strengthened each other in terms of
### looking at your feature(s) of interest?
By adding the date the contribution was received I was able to see when the 
contributions came in and how the amounts compared between candidates and
occupations.

It was noticeable that Ted Cruz began raising money before the other 
candidates. Around December of 2015, Hillary Clinton's contributions began to 
increase around December of 2015 while Bernard Sanders' contributions began to 
decline. Donald Trumps contributions began to increase around March of 2016, 
overtake all other candidates around July 2017 (except for Hillary Clinton).

It was also noticeable that retired people overtook all other occupations in the
amount contributed starting around March 2015 and continued to contribute the 
all the way up to the general election.

### Were there any interesting or surprising interactions between features?
- Ted Cruz receiving contributions in January of 2015, the next canidate
began receiving contributions around March 2015.

# Final Plots and Summary

### Plot One - Contribution Amount by Gender
```{r echo=FALSE, message=FALSE, warning=FALSE,  Plot_One}
by(tx$contb_receipt_amt, tx$gender, summary)

#Contribution Amount by Gender Boxplot
ggplot(aes(x = gender, y = contb_receipt_amt, fill = gender),
       data = tx) +
  geom_boxplot() +
  xlab('Gender') +
  ylab('Contribution Amount') +
  ggtitle('Contribution Amount by Gender Boxplot') +
  coord_cartesian(ylim = c(0, 100)) +
  scale_fill_manual(values = c('pink', 'blue')) +
  theme(legend.position = "none")
```

### Description One - Contribution Amount by Gender
I chose the Contribution Amount by Gender Box Plot because it shows there is 
a relationship between the persons gender and the contributions being made. <br>

What I discovered was: <br>
 - the median contribution of males at $50.00 was higher than females the median 
 contributions at $35.00. <br>
 - males also had a higher average contribution amount of 160.96 compared to 
 females at $119.44.  This is a difference of $41.52.<br>
 
### Plot Two - Contribution Amount by Top 5 Occupations

```{r echo=FALSE, message=FALSE, warning=FALSE,  Plot_Two}
tx.contr_occ <- summarize(occ_group, 
                              sum_occu = sum(contb_receipt_amt), 
                              mean_occu = mean(contb_receipt_amt), 
                              n = n())
tx.contr_occ <- subset(tx.contr_occ, contbr_occupation != "INFORMATION REQUESTED")
tx.contr_occ <- head(arrange(tx.contr_occ,desc(n)), n = 5)
tx.contr_occ$contbr_occupation <- ordered(tx.contr_occ$contbr_occupation, 
                                              levels = c("RETIRED", 
                                                         "NOT EMPLOYED", 
                                                         "ATTORNEY", 
                                                         "HOMEMAKER", 
                                                         "ENGINEER"))

#top five contributors by occupation
ggplot(aes(x = contbr_occupation, y = sum_occu/1000, fill = contbr_occupation), 
       data = tx.contr_occ) +
  geom_bar(stat = 'identity', color = 'darkblue') +
  xlab('Occupations') +
  ylab('Contribution Amount (Thousands)') +
  ggtitle('Contribution Amount by Top 5 Occupations') +
  scale_fill_brewer(palette = "Blues") +
  theme(legend.position = "none")
```

### Description Two - Contribution Amount by Top 5 Occupations
I chose the Contribution Amount by Top 5 Occupations
because it shows the relationship between the top five occupations.  It was 
interesting to see that retirees total contributions of over $14.5 million was
more than the the other top four occupations combined at over $9.6 million.

### Plot Three - Contribution Amount by Canidate, by Contribtution Date

```{r echo=FALSE, message=FALSE, warning=FALSE,  Plot_Three}
# Contribution Amount by Canidate, by Contribtution Date
ggplot(aes(x = contb_receipt_dt, y = sum_top_can/1000, fill = cand_nm), 
       data = tx.top_five_can) +
  geom_jitter(alpha = 1/10) +
  geom_smooth(method = 'loess') +
  scale_y_continuous(limits = c(0, 150)) +
  xlab('Date of Contribution') +
  ylab('Contribution Amount (Thousands)') +
  ggtitle('Contribution Amount by Canidate, by Contribtution Date') 
```

### Description Three - Contribution Amount by Canidate, by Contribtution Date
I chose the Contribution Amount by Canidate, by Contribtution Date because
it shows the relationship between the top five canidates that received the 
highest contribution amounts. By adding the date of contribution to the analysis 
I was able to determine how much and when the contributions were received.

Ted Cruz began raising money before the other canidates. Around December of 
2015, Hillary Clinton's contributions began to increase, and soon after 
Bernard Sanders' contributions began to decline.  Donald Trumps contributions 
began to increase around March of 2016, overtake all other canidates around July 
2017 (except for Hillary Clinton).

# Reflection

I was able to find relationships between: <br>
 - number of contributions and party affiliation <br>
 - number of contributions and presidential canidate <br>
 - number of contributions and contributor gender <br>
 - number of contributions and contributor occupation <br>
 - contributions amounts and party affiliation <br>
 - contributions amounts and presidential canidate <br>
 - contributions amounts and contributor gender 
 - contributions amounts and top five contributor occupations <br>
 - contributions amounts, top five presidential canidate, and contribution date <br>
 - contributions amounts, top five occupations, and contribution date <br>

The dataset did not contain information that I felt was usefull to get a more
detailed analysis. I wanted to know if gender played a part in the 2016 
Presidential Election, so I created a gender column by taking e contributors 
first name (contbr_first_nm), created a gender dataset (gender_df) to create 
the column.  I first tried installing and using the gender library, but found
out I needed to use the library `genderdata`.  The library genderdata is not 
available from a CRAN repository for the latest version of R. 
To install it, I used: <br>
`library(remotes)` <br>
`install_github("ropensci/genderdata")` <br>

Since there were 16,992 different occupations listed in the data set I decided 
to only analyze the top five contributor occupations.

I removed the 10,093 contributions that were greater than the $2,700 because 
they should have been refunded. 

I also removed the 12,084 contributions that were less than zero to get a better 
idea of what the contributions look liked.

## Future Work

Before I ran out of time, I wanted to look at what part of the state the 
contributions were coming from. Taking into account that Texas zipcodes ranged 
from 73301 to 88590 and wrote the following code and it is displayed in a 
chunk, but commented out:<br>
```{r}
# tx$zip = clean.zipcodes(tx$contbr_zip)
# tx$zip = substr(tx$zip, 1, 5)
# tx <- subset(tx, zip >=73301 & zip < 88590)
# data(zipcode)
# tx <- left_join(tx, zipcode, by = 'zip')
```
Hopefully, I will come back to this project and do this part of the analysis.

## Additional Resources
N/A
